A Critical Review of Sophistication and Algorithmic Model Selection

### 2014-12-19 Facticity/Facticity/paper.tex
merge
stevens gezeur weer

% in introductie?
%
% model, model class
% two-part code: sophistication is length of first part
% kolmogorov complexity?
% KC = two part vs KC in two part
% collapse, restrictions on model class (eg total instead of partial recursive, see section XX)
% conjecture

@@

Over the years one idea in particular has been proposed in many different forms. We select a description for the data by two-part coding (describing a model first, and then the data in terms of the model), and we take the complexity of only the model part as a measure of the meaningful information. If we can find a single two-part description that we can objectively call the optimal description of the data, the complexity of the data part gives us an objective property of the data. While this idea has been proposed many times in many different guises, we will, for the purposes of this paper refer to it as \emph{sophistication}\footnotemark. The general problem of how to characterize the `richness' of the information in a string rather than the plain information content (to which sophistication is but one proposed solution), we will call the `problem of interesting information'.

@@

The intuitive idea of sophistication makes one crucial demand: we can find an objectively `true' two-part description. In the framework of Komogorov complexity two observers cannot disagree by more than a constant about the information content of a dataset. In formal terms: the Kolmogorov complexity of a dataset $x$ is defined as the shortest effective description for $x$ in some Turing complete description language. This description language is usually expressed as a universal Turing machine $U(\bar\imath y) = \phi_i(y)$ where $\phi_i$ is the $i$'th partial computable function in some effective enumeration, and the $\bar\imath$ is a self-delimiting encoding of $i$. If two people disagree about the shortest effective description of $x$, we can interpret this as them using different description languages, ie. different universal Turing machines. The crucial aspect of Kolmogorov complexity, the starting theorem that makes it interesting as a function at all, is \emph{invariance}: two people using different description languages will get the same value for a string's Kolmogorov complexity, albeit up to a fixed constant. Without invariance, Kolmogorov complexity would be an arbitrary function. With invariance, we can be sure that we are talking about an intrinsic property of the data. The same consideration must be given to each proposal for formalising sophistication: is it invariant?

There are many approaches to sophistication, but the following pattern is common to all of them. We start with a space of \emph{models}. There can be finite sets, computable probability distributions or simple functions from strings to strings. The function perspective is the most generic, all other model classes can be translated to it. The approach to sophistication is to find a representation for a dataset $x$ in two parts: a computable function $f$ and and input $y$, such that $f(y) = x$. As with Kolmogorov complexity, the shorter a description is, the better. We take the shortest description and those close to it, and consider them \emph{candidates}. How close a description must be to the optimum differs between variants, it is often a pre-determined constant value. We then choose a single description from the candidate set and return its \emph{model complexity} as the sophistication. The method for determining model complexity varies. In almost all variants the candidate with the smallest model complexity determines the sophistication.

@@

In this article we will show that most proposals are not invariant. For one remaining case, we can offer no such proof, but no proof of invariance exists either. This is likely due to the complex nature of the model class chosen, and the difficulty of resolving the matter of invariance shows how complex such a sophistication measure would be to build on, even if it did prove to be invariant.

@@

We first provide a chronological overview of the variants of sophistication. In the body of the paper, we will define each variant rigorously, and in non-chronological order.

The first suggestion that a representation of a string can be separated cleanly into a structural part, containing a description of its regularities, and a noise part, containing the residual information, came from Kolmogorov himself, in a talk given at a conference in Tallinn in 1973 (first committed to paper by Cover \cite{todo} and later more extensively discussed by Vit\'anyi \cite{todo}). Here, he proposed what has become known as the \emph{Kolmogorov structure function}. Briefly, this function operates on a space of two part representations for a given string $x$, consisting of some finite set $S$ (containing $x$) and the index of $x$ in $S$. The \emph{minimal sufficient statistic} is the set with the lowest complexity for which the length of the two-part representation reaches the prefix-free Kolmogorov complexity. \pb{TODO: lowest randomness deficienty.}

Koppel \cite{} built on this notion, coining the phrase \emph{sophistication}. He changed the class of `models' from finite sets to total functions and used Levin's monotone complexity, instead of prefix-free Kolmogorov complexity. This is likely to extend the theory to infinte strings, an approach which has been abandoned by subsequent authors.

While Koppel expands the model class from finite sets to functions, he maintains a restriction that the functions be total. This is a recurring theme: if the model class is expanded to all computable functions, the the universal Turing machine $U$ itself becomes a model. If we assume that $U$ computes the function $\phi_u$ in the standard enumeration of computable functions, and that $\bar\imath y$ is a shortest description for $x$ on $U$, then $\br{u}\br{\imath}y$ is also a description for $x$. This description has a constant model size ($|\br{u}|$), ie. every $x$ has constant sophistication. This `collapse' to a universal modle is the reason that most treatments of sophistication constrain the model class to exclude the universal Turing machine. This acts as a kind of force, pushing information into the model part of the two-part code. As we shall see, defining sophistication quickly becomes a balancing act, trying to match two forces which by themselves would push information one way or the other.

In 1996 Gell-Mann and Lloyd introduced `Effective complexity'\cite{gellmann1996information}. Working from the perspective of statistical mechanics, they arrived (largely independently, it seems) at a notion that firts the mold of sophistication. Effective complexity describes a binary string in terms of a probability distribution over finite strings (an `ensemble' in the terminology of Gell-Mann and Lloyd) and the entropy of the distribution. The reasoning is that, if the string is typical for the distribution, the entropy is sufficient to describe the string given the distribution, so that the length of the description of the distribution and the size of its entropy suffices as a two-part description. The two-part description which determines the effective complexity is chosen by placing a time bound (taken from the singleton description) on the computation of the string from the distribution. The hope is that this will move deep (ie. hard to compute) information to the model part, and shallow information to the data. In a later treatment\cite{gell2004nonextensive}, Gell-Mann and Lloyd select the decisive two-part representation from the candidates based on ``the average quantities judged to be important'', which seems to be a concious choice to make the measure subjective and context-dependant (sacrificing invariance).

G\'acs, Tromp and Vit\'anyi \cite{}, built on the idea of the structure function to establish a framework of \emph{algorithmic statistics}, as an attempt to found a statistical theory not on probabilistic notions, but on combinatorial ones. This echoes Kolmogorov's own hope for the theory of information:
\begin{quotation}
\noindent Information theory must precede probability theory and not be based on it. By the very essence of this discipline, the foundations of information theory must have a finite combinatorial character. \cite{kolmogorov1983combinatorial}
\end{quotation}
Algorithmic statistics studies binary strings and their possible two-part representations. In a later paper\cite{gacs2001algorithmic}, Vit\'anyi specifically studies the role of the minimal sufficient statistic as an expression of the `meaningful' information of a string (taking the name sophistication from Koppel). Like Koppel, Vit\'anyi restricts the model class to total functions, eliminating the universal model. He balances this force by allowing the total two-part description to be non-self delimiting, so that both the representation with all information in the model part, and the representation with everything in the residual part are less efficient than the best two-part coding.\sdr{This balancing act is not yet clear to the reader}

Antunes et al. \cite{} note that the constant cutoff used to separate candidate two-part descriptions from those deemed to be too inefficient, is a source of instability. Small changes to this constant could shift the result sophistication by arbitrary amounts. They introduce a more elaborate criterion for the cutoff, calling this variant \emph{coarse sophistication}. As we shall see, this version behaves very differently from regular sophistication, and is closer to computational depth.

>>>>

% in introductie?
%
% model, model class
% two-part code: sophistication is length of first part
% KC = two part vs KC in two part 
% kolmogorov complexity?
% collapse, restrictions on model class (eg total instead of partial recursive, see section XX)
% conjecture

@@

Over the years one idea in particular has been proposed in many different forms. We select a description for the data by two-part coding (describing a model first, and then the data in terms of the model), and we take the complexity \sdr{number of bits required for} of only the model part as a measure of the meaningful information. If we can find a single two-part description that we can objectively call the optimal description of the data, the complexity of the data part gives us an objective property of the data. While this idea has been proposed many times in many different guises, \sdr{Perhaps say here already that these many attempts indicate that many people believe something can be done, and at the same time that it is not easy to get it right. Perhaps use word ``dream''} we will, for the purposes of this paper refer to it as \emph{sophistication}\footnotemark. The general problem of how to characterize the `richness' of the information in a string rather than the plain information content (to which sophistication is but one proposed solution), we will call the `problem of interesting information'.  \sdr{I'd not define it as a term here, but rather say that there are other approaches such as blah blah, but that we are not gonna talk about those in this paper. That way, we're done with it}

@@

The intuitive idea of sophistication makes one crucial demand: we can find an objectively `true' two-part description. \sdr{This is too strong, it is still okay if many descriptions are identified as long as they have more or less the same model size. Also we don't really need to make truth assumptions} In the framework of Komogorov complexity two observers cannot disagree by more than a constant about the information content of a dataset. In formal terms: the Kolmogorov complexity of a dataset $x$ is defined as the shortest effective description for $x$ in some Turing complete description language. This description language is usually expressed as a universal Turing machine $U(\bar\imath y) = \phi_i(y)$ where $\phi_i$ is the $i$'th partial computable function in some effective enumeration, and the $\bar\imath$ is a self-delimiting encoding of $i$. If two people disagree about the shortest effective description of $x$, we can interpret this as them using different description languages, ie. different universal Turing machines. The crucial aspect of Kolmogorov complexity, the starting theorem that makes it interesting as a function at all, is \emph{invariance}: two people using different description languages will get \sdr{a complexities that are within a fixed constant of each other for \emph{all} strings} the same value for a string's Kolmogorov complexity, albeit up to a fixed constant. Without invariance, Kolmogorov complexity would be an arbitrary function. With invariance, we can be sure that we are talking about an intrinsic property of the data. The same consideration must be given to each proposal for formalising sophistication: is it invariant?

There are many approaches to sophistication, but the following pattern is common to all of them. We start with a space of \emph{models}. There can be finite sets, computable probability distributions or simple functions from strings to strings. The function perspective is the most generic, all other model classes can be translated to it. The approach to sophistication \sdr{Raar begin van de zin} is to find a representation for a dataset $x$ in two parts: a computable function $f$ and and input $y$, such that $f(y) = x$. As with Kolmogorov complexity, the shorter a description is, the better. We take the shortest description and those close to it, and consider them \emph{candidates}. \sdr{We call the set of descriptions that are not much longer than the shortest one \emph{candidates}.} How close a description must be to the optimum differs between variants \sdr{varies}, it is often \sdr{but as a concrete example it could be} a pre-determined constant value. We then choose a single description from the candidate set and return its \emph{model complexity} as the sophistication. The method for determining model complexity varies. \sdr{Conceptually it is always the shortest program that implements the model, but technically this is defined in several ways, some of which are buggy. I feel we should not suggest that the notion of model complexity is very different in all approaches, because they \emph{try} to do the same thing.} In almost all variants the candidate with the smallest model complexity determines the sophistication.

@@

In this article we will show that most proposals are not invariant. \sdr{Ik wil dit nog steeds graag sterker gesteld hebben.} For one remaining case, we can offer no such proof, but no proof of invariance exists either. This is likely due to the complex nature of the model class chosen, and the difficulty of resolving the matter of invariance shows how complex such a sophistication measure would be to build on, even if it did prove to be invariant.

@@

We first provide a chronological overview of the variants of sophistication. In the body of the paper, we will define each variant rigorously \sdr{Don't emphasize distinction between formal and informal too much, it's not interesting. Everything we say should be true and clear anyway. Just say we'll provide more detailed descriptions.}, and in non-chronological order.

The first suggestion that a representation of a string can be separated cleanly into a structural part, containing a description of its regularities, and a noise part, containing the residual information, came from Kolmogorov himself \sdr{remove hint of hero worship}, in a talk given at a conference in Tallinn in 1973 (first committed to paper by Cover \cite{todo} and later more extensively discussed by Vit\'anyi \cite{todo}). Here, he proposed what has become known as the \emph{Kolmogorov structure function}. Briefly, this function operates on a space \sdr{what does ``operates on a space'' mean? It is not the domain of the function. Perhaps simplify, just say, all finite sets of binary sequences are interpreted as models} of two part representations for a given string $x$, consisting of some finite set $S$ (containing $x$) and the index of $x$ in $S$. The \emph{minimal sufficient statistic} is the set with the lowest \sdr{model?} complexity for which the length of the two-part representation reaches the prefix-free Kolmogorov complexity. \pb{TODO: lowest randomness deficienty.}

Koppel \cite{} built on this notion, coining the phrase \emph{sophistication}. He changed the class of `models' from finite sets to total functions and used Levin's monotone complexity, instead of prefix-free Kolmogorov complexity. This is likely to \sdr{This was for the purpose of extending} extend the theory to infinite strings, an approach \sdr{attempt} which has been abandoned by subsequent authors. \sdr{Dit wordt niet expliciet gemotiveerd maar is waarschijnlijk voor inf strings. In dit paper kijken we alleen naar eindige objecten}

While Koppel expands the model class from finite sets to functions, he maintains a \sdr{does impose the} restriction that the functions be total. \sdr{This collapse problem still deserves a less off-handed introduction somehow} This is a recurring theme: if the model class is expanded to all computable functions, then the universal Turing machine $U$ itself becomes a model. If we assume that $U$ computes the function $\phi_u$ in the standard enumeration of computable functions, and that $\bar\imath y$ is a shortest description for $x$ on $U$, then $\br{u}\br{\imath}y$ is also a description for $x$. This description has a constant model size ($|\br{u}|$), ie. every $x$ has constant sophistication. This `collapse' to a universal model is the reason that most treatments of sophistication constrain the model class to exclude the universal Turing machine. This acts as a kind of force \sdr{still don't like explicit force metaphor, although I do like pushing information from side to side} pushing information into the model part of the two-part code. As we shall see, defining sophistication quickly becomes \sdr{devolves into?} a balancing act, trying to match two forces which by themselves would push information one way or the other.

In 1996 Gell-Mann and Lloyd introduced `Effective complexity'\cite{gellmann1996information}. Working from the perspective of statistical mechanics, they arrived (largely independently, it seems) at a notion that fits the mold of sophistication. Effective complexity describes a binary string in terms of a probability distribution over finite strings (an `ensemble' in the terminology of Gell-Mann and Lloyd) and the entropy of the distribution. \sdr{Somehow sounds a bit unfocused} The reasoning is that, if the string is typical for the distribution, the entropy is sufficient to describe the string given the distribution, so that the length of the description of the distribution and the size of its entropy suffices as a two-part description. The two-part description that determines the effective complexity is chosen by placing a time bound (taken from the singleton description) on the computation of the string from the distribution. The hope is that this will move \sdr{just say hard to compute} deep (ie. hard to compute) information to the model part, and shallow information to the data. In a later treatment\cite{gell2004nonextensive}, Gell-Mann and Lloyd select the decisive two-part representation from the candidates based on ``the average quantities judged to be important'', which seems to be a concious choice to make the measure subjective and context-dependant (sacrificing invariance).

G\'acs, Tromp and Vit\'anyi \cite{}, built on the idea of the structure function to establish a framework of \emph{algorithmic statistics}, as an attempt to found a statistical theory not on probabilistic notions, but on combinatorial ones. This echoes Kolmogorov's own hope for the theory of information:
\begin{quotation}
\noindent Information theory must precede probability theory and not be based on it. By the very essence of this discipline, the foundations of information theory must have a finite combinatorial character. \cite{kolmogorov1983combinatorial}
\end{quotation}
Algorithmic statistics studies binary strings and their possible two-part representations. In a later paper\cite{gacs2001algorithmic}, Vit\'anyi specifically studies the role of the minimal sufficient statistic as an expression of the `meaningful' information of a string (taking the name sophistication from Koppel). Like Koppel, Vit\'anyi \sdr{et al?} restricts the model class to total functions, eliminating the universal model. He balances this force \sdr{I'm against the use of force} by allowing the total two-part description to be non-self delimiting, so that both the representation with all information in the model part, and the representation with everything in the residual part are less efficient than the best two-part coding.\sdr{Is this balancing act made sufficiently clear? I'm not happy yet about the force metaphor and explanation}

Antunes et al. \cite{} note that the constant cutoff used to separate candidate two-part descriptions from those deemed to be too inefficient, is a source of instability.\sdr{Actually their cutoff is logarithmic isn't it? Also probably very hard to understand exactly what cutoff you mean to a first time reader} Small changes to this constant could shift the resulting sophistication by arbitrary amounts. They introduce a more elaborate criterion for the cutoff, calling this variant \emph{coarse sophistication}. As we shall see, this version behaves very differently from regular sophistication, and is closer to computational depth.
